Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2012-12-29 paper.tex
more minor edits

Substituting $\{ \hat{d}_i(\theta), i \in D \}$ from~\eqref{eq:hatdfinxtheta} into~\eqref{eq:SSallparams} we obtain $SS$ minimised with respect to the data symbols,
 \begin{align*}
SS(\rho, \theta) &=\sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} \hat{d_i}(\theta) }^2 \\
&= A - \rho Z(\theta) - \rho Z^*(\theta) + L \rho^2,
\end{align*}
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta),
\]
and $Z^*(\theta)$ is the conjugate of $Z(\theta)$.  Differentiating with respect to $\rho$ and setting the resulting expression to zero gives the least squares estimator of $\rho_0$ as a function of $\theta$,
\begin{equation}\label{eq:hatrhoZ}
\hat{\rho}(\theta) = \frac{Z(\theta) + Z^*(\theta)}{2L} = \frac{1}{L}\Re(Z(\theta)),
\end{equation}
where $\Re(\cdot)$ denotes the real part.  %Since $\hat{\rho}(\theta)$ is nonnegative it follows that $\Re(Z(\theta))$ is nonnegative for all $\theta \in [-\pi, \pi)$.  
Substituting this expression into $SS(\rho, \theta)$ gives $SS$ minimised with respect to $\rho$ and the data symbols,
\[
SS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]
We again abuse notation by reusing $SS$, but this should not cause confusion as $SS(\rho,\theta)$ and $SS(\theta)$ are easily told apart by their inputs.  By definition the amplitude $\rho_0$ and its estimator $\hat{\rho}$ are positive.  However, $\hat{\rho}(\theta) = \Re(Z(\theta))$ may take negative values for some $\theta \in [-\pi,\pi)$.  %In this case we would replace $\theta$ with $\fracpart{\theta + \phi}$ and $\hat{\rho}(\theta)$ with $-\hat{\rho}(\theta)$ and observe that
%\[
%SS(\rho,\theta) = SS(-\rho, \fracpart{\theta + \pi})
%\]
The least square estimator $\hat{\theta}$ of $\theta_0$ is the minimiser of $SS(\theta)$ under the constraint $\hat{\rho}(\theta) = \Re(Z(\theta)) > 0$.  Equivalently
\[
\hat{\theta} = \arg\max_{\theta \in [-\pi,\pi)} \Re(Z(\theta))
\]
with no constraints required.

@@

\begin{lemma}\label{lem:q1k1parts}
Put
\[
q_1(\lambda) + j k_1(\lambda) = \expect\big( R_1 e^{j(\lambda + \Phi_1)} - R_1 e^{j\Phi_1}  \big).
\]
It follows that $q_1(\lambda) = o_P(1)$ and $k_1(\hat{\lambda}_L) = \hat{\lambda}_L\big(1 + o_P(1) \big)$
\end{lemma}
\begin{proof}
We have
\[
q_1(\lambda) + j k_1(\lambda) = (e^{j\lambda} - 1) \expect R_1e^{j\Phi_1} = e^{j\lambda} - 1
\]
since the mean of $R_1e^{j\Phi_1}$ is 1.  Thus $q_1(\lambda) = \cos(\lambda) - 1$ and $k_1(\lambda) = \sin(\lambda)$.  Now $q_1(\hat{\lambda}_L) = o_P(1)$ since $\hat{\lambda}_L$ converges almost surely to zero as $L\rightarrow\infty$.  By a Taylor expansion of $\sin(\lambda)$ about $\lambda = 0$ we obtain $k_1(\lambda) = \lambda \big( 1 + \zeta(\lambda) \big)$  where $\zeta(\lambda)$ is a continuous function with $\zeta(0) = 0$.  Thus $\zeta(\hat{\lambda}_L) = o_P(1)$ and $k_1(\hat{\lambda}_L) = \hat{\lambda}_L\big(1 + o_P(1) \big)$.
\end{proof}

\begin{lemma}\label{lem:q2k2parts}
Put
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big( R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big).
\]
It follows that $q_2(\hat{\lambda}_L) = o_P(1)$ and $k_2(\hat{\lambda}_L) = \hat{\lambda}_L \big( H + o_P(1) \big)$ where $H$ is defined in the statement of Theorem~\ref{thm:normality}.
\end{lemma}
\begin{proof}
Because $\hat{\lambda}_L \rightarrow 0$ almost surely as $L\rightarrow\infty$, it is only the behaviour of $q_2(\lambda)$ and $k_2(\lambda)$ around $\lambda = 0$ that is relevant.  We will examine $q_2(\lambda)$ and $k_2(\lambda)$ for $0 \leq \lambda < \tfrac{\pi}{M}$.  An analogous argument follows when $-\tfrac{\pi}{M} < \lambda < 0$.  To keep our notation clean put 
\[
\psi_k = \frac{2\pi}{M}k + \frac{\pi}{M}
\] 
with $k \in \ints$.  When $\Phi_1 \in [\psi_{k-1}, \psi_k - \lambda)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - e^{j\fracpart{\Phi_1}} &=  e^{j\big(\lambda + \Phi_1 - \tfrac{2\pi}{M}k\big)} - e^{j\big( \Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\big(\Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}},
\end{align*}
and when $\Phi_1 \in [\psi_{k} - \lambda, \psi_k)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - &e^{j\fracpart{\Phi_1}} \\
&=  e^{j\big(\lambda + \Phi_1 - \tfrac{2\pi}{M}k - \tfrac{2\pi}{M}\big)} - e^{j\big( \Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\big(\lambda - \tfrac{2\pi}{M}\big)} - 1 \big) e^{j\fracpart{\Phi_1}} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}}  + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)e^{j\fracpart{\Phi_1}}.
\end{align*}
Thus, when $\Phi_1 \in [\psi_{k-1}, \psi_k)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - e^{j\fracpart{\Phi_1}} &= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}} \\
&\hspace{0.5cm} + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)e^{j\fracpart{\Phi_1}} \chi_k(\Phi_1,\lambda)
\end{align*}
where 
\[
\chi_k(\Phi_1,\lambda) = \begin{cases}
1, & \Phi_1 \in [\psi_{k} - \lambda, \psi_k)  \\
0, & \text{otherwise}.
\end{cases}
\]
Now
\begin{align*}
q_2(\lambda) + j k_2(\lambda) &= \big(e^{j\lambda} - 1 \big) h_2(0) \\
&\hspace{0.5cm} + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big) \expect\big(R_1 e^{j\fracpart{\Phi_1}} \chi(\Phi_1,\lambda)\big)
\end{align*}
since $\expect R_1 e^{j\fracpart{\Phi_1}} = h_2(0)$ as a result of Lemma~\ref{lem:expectImRfracpart} and where $\chi(\Phi_1, \lambda) = \sum_{k \in \ints}\chi_k(\Phi_1, \lambda)$.
Now
\begin{align*}
 \expect\big(R_1 e^{j\fracpart{\Phi_1}} \chi(\Phi_1,\lambda)\big) &= \int_{0}^{2\pi}\int_{0}^{\infty} r f(r,\phi) \chi(\phi,\lambda) dr d\phi \\
&= \sum_{k\in\ints}\int_{0}^{2\pi} g(\phi) \chi_k(\phi,\lambda) d\phi \\
&= \sum_{k=0}^{M-1}\int_{\psi_k-\lambda}^{\psi_k} g(\phi)  d\phi \\
&= \lambda \sum_{k=1}^{M-1} g(\phi_k) 
\end{align*}
where for each $k$, $\phi_k \in (\psi_k-\lambda, \psi_k)$.  The third line above follows from
\end{proof}

>>>>

Substituting $\{ \hat{d}_i(\theta), i \in D \}$ from~\eqref{eq:hatdfinxtheta} into~\eqref{eq:SSallparams} we obtain $SS$ minimised with respect to the data symbols,
 \begin{align*}
SS(\rho, \theta) &=\sum_{i \in P} \abs{ y_i - \rho e^{j\theta} p_i }^2 + \sum_{i \in D} \abs{ y_i - \rho e^{j\theta} \hat{d_i}(\theta) }^2 \\
&= A - \rho Z(\theta) - \rho Z^*(\theta) + L \rho^2,
\end{align*}
where
\[
Z(\theta)  = \sum_{i \in P} y_i e^{-j\theta} p_i^* + \sum_{i \in D} y_i e^{-j\theta} \hat{d}_i^*(\theta),
\]
and $Z^*(\theta)$ is the conjugate of $Z(\theta)$.  Differentiating with respect to $\rho$ and setting the resulting expression to zero gives the least squares estimator of $\rho_0$ as a function of $\theta$,
\begin{equation}\label{eq:hatrhoZ}
\hat{\rho}(\theta) = \frac{Z(\theta) + Z^*(\theta)}{2L} = \frac{1}{L}\Re(Z(\theta)),
\end{equation}
where $\Re(\cdot)$ denotes the real part.  %Since $\hat{\rho}(\theta)$ is nonnegative it follows that $\Re(Z(\theta))$ is nonnegative for all $\theta \in [-\pi, \pi)$.  
Substituting this expression into $SS(\rho, \theta)$ gives $SS$ minimised with respect to $\rho$ and the data symbols,
\[
SS(\theta) = A - \frac{1}{L}\Re(Z(\theta))^2.
\]
We again abuse notation by reusing $SS$, but this should not cause confusion as $SS(\rho,\theta)$ and $SS(\theta)$ are easily told apart by their inputs.  By definition the amplitude $\rho_0$ and its estimator $\hat{\rho}$ are positive.  However, $\hat{\rho}(\theta) = \Re(Z(\theta))$ may take negative values for some $\theta \in [-\pi,\pi)$.  %In this case we would replace $\theta$ with $\fracpart{\theta + \phi}$ and $\hat{\rho}(\theta)$ with $-\hat{\rho}(\theta)$ and observe that
%\[
%SS(\rho,\theta) = SS(-\rho, \fracpart{\theta + \pi})
%\]
The least square estimator $\hat{\theta}$ of $\theta_0$ is the minimiser of $SS(\theta)$ under the constraint $\hat{\rho}(\theta) = \Re(Z(\theta)) > 0$.  Equivalently $\hat{\theta}$ is the maximiser of $\Re(Z(\theta))$ with no constraints required.

@@

\begin{lemma}\label{lem:q1k1parts}
Put
\[
q_1(\lambda) + j k_1(\lambda) = \expect\big[ R_1 e^{j(\lambda + \Phi_1)} - R_1 e^{j\Phi_1}  \big].
\]
We have $q_1(\lambda) = \hat{\lambda}_L o_P(1)$ and $k_1(\hat{\lambda}_L) = \hat{\lambda}_L\big(1 + o_P(1) \big)$.
\end{lemma}
\begin{proof}
We have
\[
q_1(\lambda) + j k_1(\lambda) = (e^{j\lambda} - 1) \expect R_1e^{j\Phi_1} = e^{j\lambda} - 1
\]
since the mean of $R_1e^{j\Phi_1}$ is 1.  By a Taylor expansion about $\lambda = 0$ we obtain
\[
q_1(\lambda) + j k_1(\lambda) = \lambda\big( j + \zeta(\lambda) \big)
\]
where $\zeta(\lambda)$ is a continuous function with $\zeta(0) = 0$.  Since $\hat{\lambda}_L$ converges almost surely to zero as $L\rightarrow\infty$ it follows that $\zeta(\hat{\lambda}_L) = o_P(1)$.  Thus
\[
q_1(\hat{\lambda}_L) + j k_1(\hat{\lambda}_L) = \hat{\lambda}_L\big( j + o_P(1) \big)
\]
and the lemma follow by taking real and imaginary parts.
\end{proof}

\begin{lemma}\label{lem:q2k2parts}
Put
\[
q_2(\lambda) + j k_2(\lambda) = \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big].
\]
We have $q_2(\hat{\lambda}_L) = \hat{\lambda}_L o_P(1)$ and $k_2(\hat{\lambda}_L) = \hat{\lambda}_L \big( H + o_P(1) \big)$ where $H$ is defined in the statement of Theorem~\ref{thm:normality}.
\end{lemma}
\begin{proof}
Because $\hat{\lambda}_L \rightarrow 0$ almost surely as $L\rightarrow\infty$, it is only the behaviour of $q_2(\lambda)$ and $k_2(\lambda)$ around $\lambda = 0$ that is relevant.  We will examine $q_2(\lambda)$ and $k_2(\lambda)$ for $0 \leq \lambda < \tfrac{\pi}{M}$.  An analogous argument follows when $-\tfrac{\pi}{M} < \lambda < 0$.  To keep our notation clean put 
\[
\psi_k = \frac{2\pi}{M}k + \frac{\pi}{M}
\] 
with $k \in \ints$.  When $\Phi_1 \in [\psi_{k-1}, \psi_k - \lambda)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - e^{j\fracpart{\Phi_1}} &=  e^{j\big(\lambda + \Phi_1 - \tfrac{2\pi}{M}k\big)} - e^{j\big( \Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\big(\Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}},
\end{align*}
and when $\Phi_1 \in [\psi_{k} - \lambda, \psi_k)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - &e^{j\fracpart{\Phi_1}} \\
&=  e^{j\big(\lambda + \Phi_1 - \tfrac{2\pi}{M}k - \tfrac{2\pi}{M}\big)} - e^{j\big( \Phi_1- \tfrac{2\pi}{M}k\big)} \\
&= \big(e^{j\big(\lambda - \tfrac{2\pi}{M}\big)} - 1 \big) e^{j\fracpart{\Phi_1}} \\
&= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}}  + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)e^{j\fracpart{\Phi_1}}.
\end{align*}
Thus, when $\Phi_1 \in [\psi_{k-1}, \psi_k)$,
\begin{align*}
e^{j\fracpart{\lambda + \Phi_1}} - e^{j\fracpart{\Phi_1}} &= \big(e^{j\lambda} - 1 \big) e^{j\fracpart{\Phi_1}} \\
&\hspace{0.5cm} + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)e^{j\fracpart{\Phi_1}} \chi_k(\Phi_1,\lambda)
\end{align*}
where 
\[
\chi_k(\Phi_1,\lambda) = \begin{cases}
1, & \Phi_1 \in [\psi_{k} - \lambda, \psi_k)  \\
0, & \text{otherwise}.
\end{cases}
\]
Now
\begin{align*}
q_2(\lambda) + j k_2(\lambda) &= \expect\big[ R_1 e^{j\fracpart{\lambda + \Phi_1}} - R_1 e^{j\fracpart{\Phi_1}}  \big] \\
&= \big(e^{j\lambda} - 1 \big) h_2(0) \\
&\hspace{0.5cm} + e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big) \expect R_1 e^{j\fracpart{\Phi_1}} \chi(\Phi_1,\lambda) 
\end{align*}
since $\expect R_1 e^{j\fracpart{\Phi_1}} = h_2(0) + \expect R_1 \sin\fracpart{\Phi_1} = h_2(0)$ as a result of Lemma~\ref{lem:expectImRfracpart} and where $\chi(\Phi_1, \lambda) = \sum_{k \in \ints}\chi_k(\Phi_1, \lambda)$.
Now
\begin{align*}
 \expect R_1 e^{j\fracpart{\Phi_1}} \chi(\Phi_1,\lambda)  &= \int_{0}^{2\pi}\int_{0}^{\infty} r f(r,\phi) \chi(\phi,\lambda) dr d\phi \\
&= \sum_{k\in\ints}\int_{0}^{2\pi} g(\phi) \chi_k(\phi,\lambda) d\phi \\
&= \sum_{k=0}^{M-1}\int_{\psi_k-\lambda}^{\psi_k} g(\phi)  d\phi \\
&= \lambda \sum_{k=0}^{M-1} g\big( \phi_k(\lambda) \big) 
\end{align*}
where $\phi_k(\lambda) \in (\psi_k - \lambda, \psi_k)$ for each $k$.  The third line in the previous equation follows because the $\chi_k(\phi,\lambda)$ terms inside the integral are zero for all $\phi \in [0,2\pi]$ when $k \notin \{0,\dots,M-1\}$.  We now have
\begin{equation}\label{eq:q2k2form}
q_2(\lambda) + j k_2(\lambda) = \big(e^{j\lambda} - 1 \big) h_2(0) + B(\lambda) \sum_{k=0}^{M-1} g\big(\phi_k(\lambda)\big) 
\end{equation}
where $B(\lambda) = \lambda e^{j\lambda}\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)$.  Since $\phi_k(\lambda) \rightarrow \psi_k$ as $\lambda \rightarrow 0$ and $g(\psi_k)$ is continuous at $\psi_k$ (by assumption in Theorem~\ref{thm:normality}) we have
\[
 \sum_{k=0}^{M-1} g\big(\phi_k(\hat{\lambda}_L)\big) =  \sum_{k=0}^{M-1} g(\psi_k) + o_P(1).
\]
Also $e^{j\hat{\lambda}_L} - 1 = \hat{\lambda}_L \big( j  + o_P(1) \big)$ by the argument in Lemma~\ref{lem:q1k1parts}.  Finally
\[
B(\hat{\lambda}_L) = \hat{\lambda}_L \big( o_P(1) - 2 j \sin(\tfrac{\pi}{M})\big)
\]
by Lemma~\ref{lem:Blambdaconv}.  Combining these results into~\eqref{eq:q2k2form} and taking real and imaginary parts we obtain $q_2(\hat{\lambda}_L) = \hat{\lambda}_L o_P(1)$ and 
\begin{align*}
q_2(\lambda) + &jk_2(\hat{\lambda}_L) \\
&= \hat{\lambda}_L \left( jh_2(0) - 2j\sin(\tfrac{\pi}{M}) \sum_{k=0}^{M-1} g(\psi_k) + o_P(1) \right) \\
&= \hat{\lambda}_L\big( jH + o_P(1) \big)
\end{align*}
and the lemma follows by taking real and imaginary parts.
\end{proof}

\begin{lemma}\label{lem:Blambdaconv}
$B(\hat{\lambda}_L) = \hat{\lambda}_L \big( o_P(1) - 2 j \sin(\tfrac{\pi}{M})\big)$.
\end{lemma}
\begin{proof}
A Taylor expansion about $\lambda = 0$ gives
\[
B(\lambda) = \big(  \big)\big( e^{-j\tfrac{2\pi}{M}} - 1 \big)
\]
\end{proof}
